---
title: "Tree heights"
output: pdf_document
geometry: margin=.75in
graphics: yes
header-includes:
- \usepackage{graphicx}
- \usepackage{color}
fontsize: 9pt
classoption: letter

---

```{r, setup, echo=FALSE}
library(knitr)
# Any of these options can be changed in any R chunk.
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE,
                      fig.align = "center", 
                      # Two following determine width and height
                      # of the R device on which the plots are made
                      fig.width = 7, 
                      fig.height = 6,
                      # last argument here determines the actual 
                      # width of the plot as it appears in the processed
                      # RMarkdown file 
                      out.width = "50%") 

dataDirectory <- "./Data"
```

**39 marks**


Students in two different senior statistics courses (DataViz having 107 students and EDA having 42 students) at the University of Waterloo were recently offered the opportunity to take a simple, time limited online quiz. 
Before beginning the quiz they had no idea whatsoever what questions would be asked of them, however they would each earn a single bonus mark for taking the quiz, whatever their answers.

They were asked two questions and given very little time to answer them.

The information presented in the quiz was as follows:

> *"The coast redwood is perhaps the tallest species of tree growing today.*
>
> - *Do you think the tallest tree of this species alive today is*
>
>   *A. less than XXX metres tall?*
>   *B. more than XXX metres tall?*
> 
>   *Answer A or B.*
> 
> - *Write down your best guess (in metres) of how tall you think the tallest tree might be."*

In place of *XXX* above, about half of the students (randomly selected) had the number 50 appear and the others had the number 150 appear.  The value of *XXX* presented to the students is called the *anchor* for that question.


For the record, and presumably unknown to the students taking the quiz, the tallest coast redwood tree so far found was discovered in 2006.  It was named Hyperion after the Titan of Greek mythology of that name (meaning "the high one") and was measured to be 115.7 metres tall.

The student quiz results are given in the file `trees.csv`.  
This may be read into `R` using `read.csv()` (assuming you have the csv file in a directory/folder given by `dataDirectory`) as

```{r}

trees <- read.csv(file.path(dataDirectory, "trees.csv"))
```

Note that the answers to the first question are recorded as logical values in the variable `greater` indicating whether a student believed the tallest tree to be greater than the value of the `anchor` (or not). 

Of interest are the heights the students guessed the tallest tree to be and how that compares to the true value as well as to each of the "anchor" values.

a. **(2 marks)** Describe the target population likely intended by the investigator.

   The target population likely intended by the investigator is all humans 
   who do not already know the answers of the questions asked in the quiz.

b. **(2 marks)** Describe the study population.

   The study population is all the students in two different senior statistics 
   courses at the University of Waterloo i.e., 107 students in DataViz and 42 
   students in EDA.

c. **(2 marks)** Describe the sampling plan **and** the sample it produced.
   
   Sample is self-selected because those interested in getting bonus quiz
   marks responded. 
   
   The sample produced by the sampling plan is all the students who answered the 
   quiz and earned a single bonus mark i.e., 129 out of 149 students from 
   DataViz and EDA class (93 students from DataViz and 36 students from EDA)
   

d. **(2 marks)**  How many students claimed the tallest tree would be greater than the anchor value but then guessed a smaller value?  How many claimed it would be less but then gave a greater value?

   Show your code.
   
    ```{r}
    #students that claimed the tallest tree would be greater than the anchor 
    #value but then guessed a smaller value
    greater <- trees[trees$greater == "TRUE", ]
    greater1 <- greater[greater$height < greater$anchor,]
    nrow(greater1)
    
    #students that claimed the tallest tree would be smaller than the anchor 
    #value but then guessed a greater value
    lesser <- trees[trees$greater == "FALSE", ]
    lesser1 <- lesser[lesser$height > lesser$anchor,]
    nrow(lesser1)
    
    ```
   
e. **(5 marks)** Consider the sample attribute: average `height`.  This would be the average of the student guesses for the tallest tree.  Determine the sample error of this attribute as estimate of the tallest tree for each of the following samples (show your code):

   i. *(1 mark)* All students of both classes.
   
    ```{r}
      highest_height <- 115.7 
      (sample_error_i <- mean(trees$height) - highest_height)
```
      
   ii. *(1 mark)* Students in "DataViz".
    ```{r}
       dataviz_height <- trees[trees$class == "DataViz", ]
       (sample_error_ii <- mean(dataviz_height$height)-  highest_height)
    ```
   
   
   iii. *(1 mark)* Students in "EDA".
    ```{r}
       eda_height <- trees[trees$class == "EDA", ]
       (sample_error_iii <- mean(eda_height$height) - highest_height)
    ```
      
   iv. *(1 mark)* Students who were presented with the smaller anchor.
    ```{r}
       smaller_anchor <- trees[trees$anchor == "50", ]
       (sample_error_iv <- mean(smaller_anchor$height) - highest_height)
    ```
   
   v. *(1 mark)* Students who were presented with the larger anchor.
    ```{r}
       larger_anchor <- trees[trees$anchor == "150", ]
       (sample_error_iv <- mean(larger_anchor$height) - highest_height)
    ```
   
f. **(12 marks)** In this question, you will look at producing numerical summaries for samples defined by every combination of `class` and `anchor`. 

   i. *(3 marks)* Using the `by()` function, determine the sample size and the median and inter-quartile range of the heights given by students in every combination of `class` and `anchor`.
   
      Show your code and the result it produces.
      
    ```{r}
       num_summaries <- function(samp){c("Sample Size" = length(samp),
                                         "Median" = median(samp),
                                         "Inter-quartile range" = IQR(samp))}
       (ans_by <- with(trees, by(height, list("class" = class,"anchor"=anchor),
                                 FUN=num_summaries)))
```
   ii. *(1 mark)* What is the `class`, if any, of the structure returned by `by()`?
   
    ```{r}
        (class(ans_by))
    ```
   
   
   
   iii. *(2 marks)*  Using `cut()` add a new variable to `trees` called `sizegrp` that has four groups dividing the observations according to the quartiles of the variable `height`. 
   
   
        Note, `sizegrp` should contain **no** `NA` values.
   
        Show your code and the factor levels for `trees$sizegrp`.
    ```{r}
        sizegrp <- with(trees,   cut(height, 
                        breaks = fivenum(height), # break points at min,
                                                  # quartiles, and max
                        labels = c("0-25%", "25-50%", "50-75%", "75-100%"),
                        include.lowest = TRUE, na.rm = TRUE))
                                          # lowest interval closed left
        trees$sizegrp <- sizegrp
        (levels(trees$sizegrp))
    ```
    
   iv. *(2 marks)* How large is each group in `sizegrp`?  Explain why the groups are not equal sized.  Show any code you use.
    ```{r}
       vapply(split(trees, f = sizegrp), 
                     FUN = nrow,
                     FUN.VALUE = numeric(1L))
    ```
    The groups are not equal because the height is not equally distributed among
    as they were random assumptions made by the students on the height of the 
    tallest tree.
   
   
   v. *(4 marks)*  Using `vapply()` **and** your new factor `sizegrp` produce a $3 \times 4$ matrix whose columns identify each of the 4 groups by the factor levels of `sizegrp` and whose rows are labelled  `n`, `anchor_50`, and `anchor_150`.  
   
      The first row `n` contains the sample size for each group, the second row `anchor_50` contains the proportion of students in each group presented with the anchor value of 50, and the third row `anchor_150` the proportion in each group presented with anchor value 150.   
       
      Round each proportion to 2 decimal places.
   
      Show your code.
      
      Describe any patterns you find.
    ```{r}
      sizegrp_mfunc <- function(samp){c("n" = round(nrow(samp),2),
      "anchor_50" = round(nrow(samp[samp$anchor ==     "50",])/nrow(samp),2),
      "anchor_150" = round(nrow(samp[samp$anchor ==     "150",])/nrow(samp),2))}   
       vapply(split(trees, f = sizegrp), # group by bigfarms
            FUN = sizegrp_mfunc,
            FUN.VALUE = c("n" = 0, "anchor_50" = 0, "anchor_150" = 0))
```
     The proportion in anchor_50 decreases as the quartile increases from 0-25% 
     to 75-100% (0-25% has the highest proportion while 75-100% has the lowest 
     proportion) while the proportion in anchor_150 increases as the quartile 
     increases from 0-25% to 75-100% (0-25% has the lowest proportion while 
     75-100% has the highest proportion)
     
     
g. **(11 marks)** Here we will look at the entire distribution of height provided by students in different groups.

   i. *(2 marks)*  Using bin widths of 50, draw a histogram of the heights provided by all students.   Add a "blue" vertical dashed line of width 3 at the height obtained by Hyperion. 
   
      Show your code.
      
    ```{r}
      xlim <- extendrange(trees$height)
      hist(trees$height, main = "Distribution of heights provided by students",
         xlim = xlim, xlab = "Height", breaks = 50)
      abline(v=115.7, col = "blue", lty = 2, lwd = 3)
    ```
   
   ii. *(5 marks)* In this question, you are required to use  `sapply()` with `split()` 
       to draw a histogram for each `anchor` **and** to return (as the value of the `sapply()`) 
       the average heights for each anchor.
       
       The histograms are to be drawn subject to the following constraints:
       
       - the two histograms appear side by side in the same display 
       
       - each histogram has the same x range on its horizontal axis
       
       - each histogram uses bins of width 50
       
       - each histogram has a vertical blue line at Hyperion's height
       
       - each histogram has a vertical red line at the value of its anchor
       
       - the title of each histogram says the anchor for which it is displaying values
       
       - label the axes as appropriate
       
       - on the right hand plot **only**, add a legend to indicate which vertical line is the anchor value and which is Hyperion's height
       
       Note: In your `RMarkdown` you might want to use the following header for the R chunk
       
       `{r, out.width = "100%", fig.width = 10, fig.height = 4}`
       
       Show your code.
       
    ```{r, out.width = "100%", fig.width = 10, fig.height = 4}
       (hist_sapply <- sapply(split(trees$height, f = trees$anchor), # group by bigfarms
                     FUN = mean))
       hist_split <- (split(trees$height, f = trees$anchor))
       savePar <- par(mfrow = c(1,2))
       xlim <- extendrange(trees$height)
       hist(hist_split[[1]], main = "Height distribution for anchor = 50",
         xlim = xlim, xlab = "Height", w = 50)
      abline(v=115.7, col = "blue", lty = 2, lwd = 2)
      abline(v=50, col = "red", lty = 2, lwd = 2)
      hist(hist_split[[2]], main = "Height distribution for anchor = 150",
         xlim = xlim, xlab = "Height", w = 50)
      abline(v=115.7, col = "blue", lty = 2, lwd = 2)
      abline(v=150, col = "red", lty = 2, lwd = 2)
      legend("topright",  bty = "n",
       legend = c(expression("anchor value"), 
                  expression("Hyperion's height"),
                  ""),
       col = c("red", "blue", "white"), 
       lty = c(2, 2, 3), lwd = 3)
       par(savePar)

    ```
       
      
   iii. *(4 marks)* **Overlaid quantiles plot**  On a single plot, for each `anchor` draw overlaid the sample quantiles of the heights for that anchor.  When drawing the plot,
   
        - make sure all heights fit on the plot
        - label the axes appropriately
        - title the plot "Comparing anchors using quantiles"
        - use arguments `pch = 19, type = "b"`
        - distinguish the two anchor groups by colour
        - add a legend to identify the two anchor groups
        
        Show your code.
        
    ```{r}
        ylim <- extendrange(trees$height)
        plot(x = ppoints(length(hist_split[[2]])),
             y = sort(hist_split[[2]]),
             main = "Comparing anchors using quantile",
             ylab = "height", xlab = "p",
             pch = 19, type = "b", ylim = ylim,
             col = adjustcolor("steelblue", alpha.f = 0.5))
         points(x = ppoints(length(hist_split[[1]])),
                y = sort(hist_split[[1]]),
        pch = 19, type = "b", col = adjustcolor("firebrick", alpha.f = 0.5))
     legend("topleft", legend = c("anchor = 50", "anchor = 150"),
      col = adjustcolor(c("firebrick", "steelblue"), alpha.f = 0.5), pch = 19)
    ```
h. **(3 marks)** **Conclusions**:  Based on your entire analysis above, what do you conclude from this study?


    There seems to be outliers for anchor value 150, where students have input 
    1000 and 900 for the height which is very large compared to the real height 
    of 115.7. The other heights for anchor value 150 seem to be between 
    approximately 50 to 400 with the majority between 100-200 which is closer
    to 115.7. For anchor value 50, there are no outliners and most students
    have inputted values in the range 0-200 which seems to be more closer to 
    true value compared to the values inputted for anchor value 150. We can 
    conclude that having the anchor value as 50, helped the students give a 
    more accurate estimate of the height of the tree compared to anchor value 
    150.
    
    ```{r}
    ?hist
    ```
    
   